---
title: Mini Data Analysis with datateachr
author: Eric Liu
date: October 9, 2021
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = 'left', echo = TRUE)
```

## 0.1 Introduction

This is a data exploration into the datasets from the [datateachr](https://github.com/UBC-MDS/datateachr) package. Initially, 4 of the 7 available datasets will be chosen to be further explored and for which a proposed research question will be formulated. We will then choose one data set for an in depth analysis. 

The primary tool for data cleaning and visualization will be *dplyr* and *ggplot* respectively.

## 0.2 Load required libraries
```{r libraries, message= FALSE}
library(datateachr)
library(tidyverse)
library(readr)
```

## 1.1 Chosen datasets
1. apt_buildings
2. cancer_sample
3. MSK-IMPACT Clinical Sequencing Cohort, available [here](https://www.cbioportal.org/study/clinicalData?id=msk_impact_2017)
4. buparlisib and letrozole in ER+ metastatic breast cancer, available [here](https://www.cbioportal.org/study/clinicalData?id=brca_mskcc_2019)

## 1.2 Data overview 
The first two data sets can simply be loaded natively in R using `apt_buildings` and `cancer_sample` as we have loaded the datateachr package. The last two datasets must be first downloaded and imported using the readr package. 
```{r explore, message = FALSE}
#function to quickly look at all 4 datasets in the same manner
quick_look <- function(df) {
  glimpse(df)
  print(class(df))
  paste("Contains NA?", any(is.na(df)))
}

#first we will look at the apt_buildings dataset from datateachr
apt_buildings <- apt_buildings
quick_look(apt_buildings)

#repeated for cancer_sample 
cancer_sample <- cancer_sample
quick_look(cancer_sample)

#we must first import the last 2 data sets, and again look at their features
msk <- read_tsv("msk_impact_2017_clinical_data.tsv")
quick_look(msk)
metastatic_BC <- read_tsv("brca_mskcc_2019_clinical_data.tsv")
quick_look(metastatic_BC)
```

## 1.3 Data exploration 
We will further explore the 4 datasets and formulate research questions for each.

