---
title: "Mini Data Analysis Milestone 2"
output: md_document
---

*To complete this milestone, you can either edit [this `.rmd` file](https://raw.githubusercontent.com/UBC-STAT/stat545.stat.ubc.ca/master/content/mini-project/mini-project-2.Rmd) directly. Fill in the sections that are commented out with `<!--- start your work here--->`. When you are done, make sure to knit to an `.md` file by changing the output in the YAML header to `github_document`, before submitting a tagged release on canvas.*

# Welcome back to your mini data analysis project!

This time, we will explore more in depth the concept of *tidy data*, and hopefully investigate further into your research questions that you defined in milestone 1.

**NOTE**: The main purpose of the mini data analysis is to integrate what you learn in class in an analysis. Although each milestone provides a framework for you to conduct your analysis, it's possible that you might find the instructions too rigid for your data set. If this is the case, you may deviate from the instructions -- just make sure you're demonstrating a wide range of tools and techniques taught in this class.

Begin by loading your data and the tidyverse package below: 

```{r setup, message = FALSE}
library(datateachr) # <- might contain the data you picked!
library(tidyverse)
```

```{r data-loading, message = FALSE}
msk <- msk <- read_tsv("https://raw.githubusercontent.com/stat545ubc-2021/mini-data-analysis-EL/main/data/msk_impact_2017_clinical_data.tsv")

#external datasets are first downloaded and loaded using readr
```

# Learning Objectives

By the end of this milestone, you should:

+ Become familiar with manipulating and summarizing your data in tibbles using `dplyr` and `tidyr`, with a research question in mind.
+ Understand what *tidy* data is, and how to create it. In milestone 3, we will explore when this might be useful.
+ Generate a reproducible and clear report using R Markdown.
+ Gain a greater understanding of how to use R to answer research questions about your data.

**Things to keep in mind**

+ Remember to document your code, be explicit about what you are doing, and write notes in this markdown document when you feel that context is required. Create your analysis as if someone else will be reading it! **There will be 2.5 points reserved for reproducibility, readability, and repo organization.**

+ Before working on each task, you should always keep in mind the specific **research question** that you're trying to answer.

# Task 1: Process and summarize your data (15 points)

From milestone 1, you should have an idea of the basic structure of your dataset (e.g. number of rows and columns, class types, etc.). Here, we will start investigating your data more in-depth using various data manipulation functions. 

### 1.1 (2.5 points) 

First, write out the 4 research questions you defined in milestone 1 were. This will guide your work through milestone 2:

<!-------------------------- Start your work below ---------------------------->
1. Which `Cancer Type` is the most deadly?
2. Are patient variables or sample variables more predictive of overall survival?
3. What are the differences in the genomic and mutational variables between primary and Metastasis samples?
4. What are the genomic and mutational consequences of a patient's smoking history?
<!----------------------------------------------------------------------------->

### 1.2 (10 points)

Now, for each of your four research questions, choose one task from options 1-4 (summarizing), and one other task from 4-8 (graphing). You should have 2 tasks done for each research question (8 total). Make sure it makes sense to do them! (e.g. don't use a numerical variables for a task that needs a categorical variable.). Comment on why each task helps (or doesn't!) answer the corresponding research question. 

Ensure that the output of each operation is printed! 

**Summarizing:**

1. Compute the *range*, *mean*, and *two other summary statistics* of **one numerical variable** across the groups of **one categorical variable** from your data.
2. Compute the number of observations for at least one of your categorical variables. Do not use the function `table()`!
3. Create a categorical variable with 3 or more groups from an existing numerical variable. You can use this new variable in the other tasks! *An example: age in years into "child, teen, adult, senior".*
4. Based on two categorical variables, calculate two summary statistics of your choosing.

**Graphing:**

5. Create a graph out of summarized variables that has at least two geom layers.
6. Create a graph of your choosing, make one of the axes logarithmic, and format the axes labels so that they are "pretty" or easier to read. 
7. Make a graph where it makes sense to customize the alpha transparency.
8. Create 3 histograms out of summarized variables, with each histogram having different sized bins. Pick the "best" one and explain why it is the best.

Make sure it's clear what research question you are doing each operation for!

<!------------------------- Start your work below ----------------------------->
**1. Which `Cancer Type` is the most deadly?**

While overall survival will also reflect other factors such as treatment and stage of the disease, it will be the best surrogate for the disease burden of each cancer type. 

```{r question1a}
# Calculate summary statistics of overall survival for each cancer type
most_deadly <- msk %>%
  filter(!is.na(`Overall Survival (Months)`)) %>%
  group_by(`Cancer Type`) %>%
  summarize(median_os = median(`Overall Survival (Months)`),
            iqr_os = IQR(`Overall Survival (Months)`),
            mean_os = mean(`Overall Survival (Months)`),
            range_os = diff(range(`Overall Survival (Months)`)),
            n_samples = n()) %>%
  arrange(median_os)

head(most_deadly, 20)

# store the 10 most deadly types for graphing
most_deadly_types <- most_deadly %>%
  head(10) %>%
  select(`Cancer Type`) %>%
  as_vector()
```
We can see that 9 cancer types have a median overall survival of under a year. Generally, these were cancers with a small number of samples, of unknown primary, or pancreatic and lung cancer. This can be represented graphically with box plots and dot plots that show the overall survival for each sample, segregated by `Cancer Type`.

```{r question1b}
# We filter the msk dataset to include only the 10 most deadly cancer types defined previously
# We created a new factor variable for cancer type that is reordered from lowest to highest median overall survival
# We then generate boxplots and violin plots of the overall survival for the 10 cancer types
msk %>%
  filter(`Cancer Type` %in% most_deadly_types & !is.na(`Overall Survival (Months)`)) %>%
  mutate(type = fct_reorder(`Cancer Type`, `Overall Survival (Months)`)) %>%
  ggplot(aes(x= type, y = `Overall Survival (Months)`, color = type)) +
  geom_violin(trim=FALSE) +
  geom_boxplot(width = 0.1, outlier.color = "black", outlier.size = 0.5) +
  theme(axis.text.x = element_text(size = 6, angle = 45, vjust = 1, hjust = 1), legend.position = "none")

# changes to the theme makes the geoms more visually appealing, the axis labels more readable, and the outliers more obvious. 

```
We visualized the boxplots for the 10 cancer types with the shortest median overall survival. While Violin plots helped us to peek at the underlying sample distributions, we must also consider the bias from the variable sample size of each type.

**2. Are patient variables or sample variables more predictive of overall survival?**

We will look at the following patient variables: `Smoking History` and `Sex`. We will look at the following sample variables: `Specimen Type` and `Specimen Preservation Type`.

```{r question2a, message=FALSE}
# remove NA values in overall survival and group by chosen patient variables to compute summary statistics
msk_patient <- msk %>%
  filter(!is.na(`Overall Survival (Months)`)) %>%
  group_by(`Smoking History`, `Sex`) %>%
  summarize(median_os = median(`Overall Survival (Months)`),
            mean_os = mean(`Overall Survival (Months)`),
            count = n())

print(msk_patient)

# repeated for chosen sample variables
msk_sample <- msk %>%
  filter(!is.na(`Overall Survival (Months)`) & `Specimen Type` != "#N/A") %>%
  group_by(`Specimen Type`, `Specimen Preservation Type`) %>%
  summarize(median_os = median(`Overall Survival (Months)`),
            mean_os = mean(`Overall Survival (Months)`),
            count = n())

print(msk_sample)
```

From the calculated summary statistics, it appears that `Smoking History` is associated with overall survival while `Sex` is not. For sample variables, it appears that both the variables we consider may be associated with overall survival. We will then generate a scatter plot of `Fraction Genome Altered` and `Sample coverage`, coloring the inidividual points patient survival status. 

```{r question2b}
# NA's in overall survival are removed, before a point geom is generated
msk %>%
  filter(!is.na(`Overall Survival (Months)`) & !is.na(`Fraction Genome Altered`)) %>%
    ggplot(aes(x = `Fraction Genome Altered`, y = `Sample coverage`, color = `Overall Survival Status`)) +
  geom_point(alpha = 0.25, size = 1)
#alpha is changed to uncover overlapping points
```

From the plot, it appears that sample with higher coverage were typically from living patients.

**3. What are the genomic and mutational differences between primary and Metastasis samples?**

We begin by calculating summary statistics for metastatic and primary samples. 

```{r question3a}
# NA's in fraction genome altered is removed before summary statistics are calculated for each sample type
msk_sample_type <- msk %>%
  filter(!is.na(`Fraction Genome Altered`)) %>%
  group_by(`Sample Type`) %>%
  summarize(median_genome_altered = median(`Fraction Genome Altered`),
            count = n(),
            iqr_genome_altered = IQR(`Fraction Genome Altered`),
            mean_genome_altered= mean(`Fraction Genome Altered`),
            mean_mutation_count = mean(`Mutation Count`),
            median_mutation_count = median(`Mutation Count`))


print(msk_sample_type)
```
We then plot the density curves of fraction of genome altered for both types of samples. From our summarize data, we can also plot a vertical line representing the median for both groups. 

```{r question3b}
# We remove rows that contains NA values for the fraction genome altered variable
# A density curve is plotted for each sample type 
msk %>%
  filter(!is.na(`Fraction Genome Altered`)) %>%
  ggplot(aes(x = `Fraction Genome Altered`, fill = `Sample Type`)) +
  geom_density(aes(color = `Sample Type`), alpha = 0.25) +
  geom_vline(data=msk_sample_type, aes(xintercept=median_genome_altered, color=`Sample Type`),
             linetype="dashed")

#we can add the median as a vertical line using our summarize data
```

This illustrates the substantial difference in `Fraction Genome Altered` between the two samples types. 

**4. What are the genomic and mutational consequences of a patient's smoking history?**

We do a similar analysis as for question 3, but this time grouping the data by smoking history. We observe that the differences in the calculated summary statistics between groups are small. 

```{r question4a}
# NA's in fraction genome altered and smoking history are removed
# Summary statistics are calculated for each category of smoking history
msk_smoking_history <- msk %>%
  filter(!is.na(`Fraction Genome Altered`) & !is.na(`Smoking History`)) %>%
  group_by(`Smoking History`) %>%
  summarize(median_genome_altered = median(`Fraction Genome Altered`),
            count = n(),
            iqr_genome_altered = IQR(`Fraction Genome Altered`),
            mean_genome_altered= mean(`Fraction Genome Altered`),
            median_mutation_count = median(`Mutation Count`),
            mean_mutation_count = mean(`Mutation Count`))

print(msk_smoking_history)
```
This is then visualized and can be contrasted with the plot generated for question 3. It is clear that any differences in `Fraction Genome Altered` between smoking history groups is smaller than between sample types. 

```{r question4b}
# NA's in the variables to be plotted are removed
# A density curve is plotted for each smoking history
msk %>%
  filter(!is.na(`Smoking History`) & !is.na(`Fraction Genome Altered`)) %>%
  ggplot(aes(x = `Fraction Genome Altered`, fill = `Smoking History`)) +
  geom_density(aes(color = `Smoking History`), alpha = 0.25) +
  geom_vline(data=msk_smoking_history, aes(xintercept = median_genome_altered, color = `Smoking History`),
             linetype="dashed")
```
<!----------------------------------------------------------------------------->

### 1.3 (2.5 points)

Based on the operations that you've completed, how much closer are you to answering your research questions? Think about what aspects of your research questions remain unclear. Can your research questions be refined, now that you've investigated your data a bit more? Which research questions are yielding interesting results?

<!------------------------- Write your answer here ---------------------------->
**1. Which `Cancer Type` is the most deadly?**

From our analysis, we were able to arrange the cancer types in order of lowest median overall survival, and this showed us a few interesting observations. First, 5 of the 10 most deadly cancer types identified in this manner had a sample size of under 10 out of over `nrow(msk)` samples. As well, since Cancer of Unknown Primary was the third deadliest cancer, we might be persuaded to make the conclusion that disease in which we have less knowledge will lead to a lower patient life expectancy. However, we must also consider that such observations are more likely attributed to the small sample sizes for certain types and that cancers of unknown primary would likely be late stage disease that has metastasized, making it more deadly than disease that is detected earlier or has yet to metastasize. 

**2. Are patient variables or sample variables more predictive of overall survival?**

In our analysis, we only looked at two patient variables and two sample variables. While this may have revealed the association between certain variables and overall survival, this was inherently biased as they were chosen and not all possible variables were considered. Unfortunately, the analysis done was of extremely limited scope, and fails to make full use of the dataset. It remains an open question whether patient or sample variables are more predictive of overall survival, and we may have to account for the inherent bias present in the dataset as it relates to this question (e.g. uneven number of patient and sample variables).

**3. What are the differences in the genomic and mutational variables between primary and Metastasis samples?**

From out dataset, we had nearly equal rows of metastatic and primary samples. Looking at the mean and median of the fraction of the genome altered in these two types of samples, it is evident that it was generally higher in the metastatic samples. This was confirmed when we plotted the density curves of the fraction of the genome altered for these two groups, as the distribution of the metastatic samples was shifted to the right. However, this was not reflected in the mean and median of `Mutation Count` when stratified by sample type, which may weaken the conclusions that are drawn from looking at only `Fraction Genome Altered`. 

**4. What are the genomic and mutational consequences of a patient's smoking history?**

We were able to calculate summary statistics for `Fraction Genome Altered` and `Mutation Count`, but differences were small. This was also reflected in the density curves of `Fraction Genome Altered`grouped by `Smoking History`, which overlapped significantly. In our analysis, we are only able to show the association between a patient's smoking history and the two variables that we used as surrogate for the genomic and mutation characteristics of the disease. While the difference in the mean `Mutation Count` between smokers and non-smokers is intriguing, it may be due to differences in the types of cancer that smoking is a risk factor for rather than due to direct effect of smoking on the DNA of the tumour.

<!----------------------------------------------------------------------------->

# Task 2: Tidy your data (12.5 points)

In this task, we will do several exercises to reshape our data. The goal here is to understand how to do this reshaping with the `tidyr` package.

A reminder of the definition of *tidy* data:

- Each row is an **observation**
- Each column is a **variable**
- Each cell is a **value**

*Tidy'ing* data is sometimes necessary because it can simplify computation. Other times it can be nice to organize data so that it can be easier to understand when read manually. 

### 2.1 (2.5 points)

Based on the definition above, can you identify if your data is tidy or untidy? Go through all your columns, or if you have >8 variables, just pick 8, and explain whether the data is untidy or tidy.

<!--------------------------- Start your work below --------------------------->
In our dataset, each row corresponds to a unique tumor sample from patients in the MSK-IMPACT Clinical Sequencing Cohort. As well, each column is a variable that describes either a characteristic of the patient (e.g. `Patient ID`, and `Smoking History`) or of the sample itself (e.g. `Cancer Type`, `Oncotree Code`, `Fraction Genome Altered`, `Metastatic Site`, `Primary Tumor Site`, and `Sample Type`). The data is structured in a manner in which each cell only contains a single value, but in some cases it is missing. Therefore, the structure of this dataset is tidy, despite it being incomplete in the sense that there are missing values.

```{r tidydata}
# we select 8 variables from the original dataset to illustrate the data is clean
msk_tidy <- msk %>%
  select(`Patient ID`, `Cancer Type`, `Oncotree Code`, `Fraction Genome Altered`, `Metastatic Site`, `Primary Tumor Site`, `Smoking History`, `Sample Type`)

head(msk_tidy, 10)
```
<!----------------------------------------------------------------------------->

### 2.2 (5 points)

Now, if your data is tidy, untidy it! Then, tidy it back to it's original state.

If your data is untidy, then tidy it! Then, untidy it back to it's original state.

Be sure to explain your reasoning for this task. Show us the "before" and "after".

<!--------------------------- Start your work below --------------------------->
We will begin be selecting 8 variables of the original `ncol(msk)`, and untidying the data so that each row corresponds to a patient by combining the `Fraction Genome Altered` of primary and metastatic samples. In the original dataset, each observation corresponded to a different sample, which may have been from the same patient. Now, if there are multiple samples from the same patient, they will no longer be individual rows but rather as multiple observations contained in the same row. 

```{r untidying, warning=FALSE}
# we use pivot wider to reduce the number of rows from the initial data set
# Sample type is used for the names of the new columns
# Fraction of genome altered becomes the values for the new variables

msk_untidy <- msk_tidy %>%
  pivot_wider(id_cols = c(-`Sample Type`, -`Fraction Genome Altered`), 
              names_from = `Sample Type`,
              values_from = `Fraction Genome Altered`)

head(msk_untidy, 10)
```
Our data is now untidy, as we have a variable, `Sample Type`, that is in the header of the table rather than as a column. As well, since some patients had multiple samples of the same `Sample Type`, some cells contain more than 1 value. Thus, the data has been untidied. We will now tidy the data back to its original state, so that Sample Type is again its own variable, and each row corresponds to a unique sample. 

```{r retidy}
# restore the Sample Type Column, and combine the values from the two variables in the untidy dataset into a single variable
msk_retidy <- msk_untidy %>% 
  pivot_longer(cols = c(Primary, Metastasis),
               names_to = "Sample Type",
               values_to = "Fraction Genome Altered") %>%
  unnest()
# we must use unnest to regenerate each of the orignal columns as the pivot longer function initially creates a list column
# rows that contained two values stored as a vector of length 2 in the list column are now individual samples represented in their own row

head(msk_retidy, 10)
``` 
<!----------------------------------------------------------------------------->

### 2.3 (5 points)

Now, you should be more familiar with your data, and also have made progress in answering your research questions. Based on your interest, and your analyses, pick 2 of the 4 research questions to continue your analysis in milestone 3, and explain your decision. 

Try to choose a version of your data that you think will be appropriate to answer these 2 questions in milestone 3. Use between 4 and 8 functions that we've covered so far (i.e. by filtering, cleaning, tidy'ing, dropping irrelvant columns, etc.). 

<!--------------------------- Start your work below --------------------------->
From our initial 4 research questions, 2 have been selected for the remainder of this project.

**Are patient variables or sample variables more predictive of overall survival?**

While our data analysis and visualization has helped us in exploring this question, it was unable to fully address the scope required to provide a more comprehensive conclusion. As well, I believe this would be a question that is better addressed by modelling rather than data visualization, so I believe it well lend itself well to the next steps of the project.

**What are the differences in the genomic and mutational variables between primary and Metastasis samples?**

Our analysis and visualization has already revealed very intriguing differences between primary and metastasis samples. In the subsequent steps of this project, these observations can be fully dissected. Metastasis is also a relevant clinical outcome that may be predictive of disease burden, it would be interesting to see if this is independent of other changes associated with metastatic disease.  

We will now subset our data in preparation for the next steps of this project.

```{r subset}
# we begin by selecting variables that are relevant to our research questions, and filtering for NA values as necessary.
# we combine NA in smoking history with the unknown category, as they effectively have the same meaning
# we arrange by overall survival to make the dataframe more intuitive for ourselves
msk_clean <- msk %>%
  select(-`Study ID`, -`Sample ID`, -`Cancer Type Detailed`, -`Matched Status`, -`Sample Class`, -`Sample Collection Source`) %>%
  filter(!is.na(`Overall Survival (Months)`)) %>%
  mutate(`Smoking History` = ifelse(is.na(`Smoking History`), Unknown, `Smoking History`)) %>%
  arrange(desc(`Overall Survival (Months)`))

head(msk_clean, 10)
```
<!----------------------------------------------------------------------------->

*When you are done, knit an `md` file. This is what we will mark! Make sure to open it and check that everything has knitted correctly before submitting your tagged release.*

### Attribution

Thanks to Victor Yuan for mostly putting this together. 