---
title: Mini Data Analysis Project With datateachr
author: Eric Liu
date: October 9, 2021
output: md_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = 'left', echo = TRUE)
# set working directory to be parent folder as it contains the external datasets
```

# 0.1 Introduction

This is a data exploration into the datasets from the [datateachr](https://github.com/UBC-MDS/datateachr) package as well as external datasets from [cBioPortal](https://www.cbioportal.org). Initially, 4 datasets will be chosen to be further explored. We will then choose a single dataset for which we will formulate research questions to address with an in-depth analysis.

The primary tool for data cleaning and visualization will be *dplyr* and *ggplot* respectively.

# 0.2 Load required libraries
```{r libraries, message= FALSE}
#packages must first be installed locally before they can be loaded
library(datateachr)
library(tidyverse)
library(readr)
```

# 1.1 Choosing datasets
I have chosen two datasets from *datateachr* and two external datasets that were of personal interest to me as the starting point for our analysis.

1. apt_buildings
2. cancer_sample
3. MSK-IMPACT Clinical Sequencing Cohort, available [here](https://www.cbioportal.org/study/clinicalData?id=msk_impact_2017). This contains clinical data on patients with various types of cancer who had their tumour sequenced. 
4. buparlisib and letrozole in ER+ metastatic breast cancer, available [here](https://www.cbioportal.org/study/clinicalData?id=brca_mskcc_2019). This contains clinical data on patients with cancer who were enrolled into a clinical trial for the two two drugs.

# 1.2 Data overview 
We will now use base R and tidyverse tools to gain a better understanding of the data contained within each dataset. The first two datasets can simply be loaded natively in R using `apt_buildings` and `cancer_sample` as we have already loaded the *datateachr* package. The last two datasets must be first downloaded and then imported using the *readr* package. Using the `quick_look` function, we will be able to extract the dimensions, name of the variables, and class of the dataset, as well as whether it contains any NA values. 
```{r explore, message = FALSE}
#function to quickly gain key insights into a dataset
quick_look <- function(df) {
  glimpse(df)
  print(class(df))
  paste("Contains NA?", any(is.na(df)))
}

#first we will look at the apt_buildings dataset from datateachr
apt_buildings <- apt_buildings
quick_look(apt_buildings)

#repeat for cancer_sample 
cancer_sample <- cancer_sample
quick_look(cancer_sample)

#we must first import the last 2 data sets, and again look at their features
msk_online <- 

msk <- read_tsv("https://raw.githubusercontent.com/stat545ubc-2021/mini-data-analysis-EL/main/data/msk_impact_2017_clinical_data.tsv")
metastatic_BC <- read_tsv("https://raw.githubusercontent.com/stat545ubc-2021/mini-data-analysis-EL/main/data/brca_mskcc_2019_clinical_data.tsv")
quick_look(metastatic_BC)

#keept the msk variable names as strings as they could be easily incorporated into markdown text.  
```

# 1.3 Choosing 2 of the 4 datasets
The above data exploration has provided us with basic but valuable insights into the data contained within each dataset. All datasets followed the "tibble" convention of having columns as variables and rows as individual observation. Data in this form is clean when it does not contain any missing values.

1. The `apt_buildings` dataset is composed of up to 37 aspects of 3,455 buildings in Toronto. As the data contains NA values, it is not clean and hence why not all 37 aspects will be described for each building. It appears to be primarily categorical data, as many of the numeric variables are not continuous, but rather only take on a subset of discrete values (ie. `no_of_elevators` only contains 11 distinct values).

2. The `cancer_sample` dataset is composed of 32 features of 569 images of cancer samples. As the data does not contain NA values, it is clean. it is composed of only continuous data, other than `diagnosis` which is categorical and `ID` which uniquely identifies each row.

3. The `msk` dataset is composed of 26 aspects of 10,945 patient enrollments into the MSK-IMPACT tumour sequencing study. As the data contains NA values, it is not clean. It contains variables describing characteristics of both the patient and the tumour. 

4. The `metastatic_BC` dataset is composed of 29 aspects of 70 patient enrollments into a clincial trial for the drugs buparlisib and letrozole in estrogen receptor-positive, metastatic breast cancer. As the data contains NA values, it is not clean.Variables describe both patient and tumour characteristics, with several variables related to relevant aspects and outcomes of the clinical trial. 

From the insights into the dataset that we extracted, I have chosen the `msk` and `metastatic_BC` datasets as they are the only two containing clinical data of cancer patients. This is an area I am passionate about and I hope to pursue a career as an oncologist or translational scientist in the future. As well, while the `cancer_sample` dataset is also related to the field, it contains data derived from machine learning deconvolution of cancer sample images rather than clinical data of the patients themselves. Hence, the 2 external datasets better reflect my personal interests and were chosen. 

# 1.4 Choosing a single dataset
Looking at the `msk` dataset, there are several variables that describe a potential outcome of interest, such as `Overall Survival (Months)` or `Metastatic Site`. A potential research question could be "How is `Smoking History` related to the patient's disease course?"

Looking at the `metastatic_BC` dataset, it appears a variable describing an outcome of interest could be `Best Response to Therapy`. Particularly, a potential research question could be "Is the mutational and genomic characteristics of a patient's tumor predictive of their `Best Response to Therapy`?". 

After having formulated research questions for both the datasets, I have chosen the `msk` dataset as it is a larger cohort of patients and includes various types of cancer. On the other hand, the `metastatic_BC` dataset was gather from a clinical study and thus represents a more homogeneous group of patients. The data was also gathered with the hypothesis of testing the effectiveness of the experimental intervention in mind. Thus, I believe the diversity of the `msk` dataset might lend to a more interesting and open-ended exploration, hence why I chose it for the next section of the project. 

# 2.1 Data exploration 
I will perform 4 data analysis exercises to gain a deeper understanding of the data contained in `msk`. 

**Exercise 1 :** Filtering of dataset to include only the 4 most common types of cancer. This was done to make the dataset more manageable to work with, as well as to stratify by cancer type in our analysis while maintaining readability.
``` {r exercise-1}
# we first summarize the dataset to reflect how many patients there are for each cancer type, before selecting the top 4 and storing them in a vector
 most_common_cancer_types <- msk %>%
  group_by(`Cancer Type`) %>%
  summarize(n()) %>%
  arrange(desc(`n()`)) %>%
  head(4) %>%
  select(`Cancer Type`) %>%
  as_vector()

#we then use the vector of the top 4 most common cancer types to filter the entire dataset 
msk_most_common <- msk %>% filter(`Cancer Type` %in% most_common_cancer_types)
```

**Exercise 2 :** Plot of the density of `Mutation Count`. Since we took a subset from the initial dataset that only contains the 4 most common cancer types, we can observe the distribution for each type (whereas including all cancer types from the orignal dataset would overwhelm the plot). 
``` {r exercise-2}
#To plot a single continuous variable, only the x aesthetic needs to be specified. It is plotted on a log scale to get a better sense of the distribution of the variable. 
msk_most_common %>% ggplot(aes(x = `Mutation Count`)) +
  geom_density(aes(fill = `Cancer Type`), alpha = 0.3) +
  scale_x_log10() +
  scale_color_discrete()
#The mutation count distribution is plotted for each cancer type so alpha is adjusted to increase the readability of the plot
``` 

**Exercise 3 :** Plot of the distribution of `Fraction Genome Altered`. Histograms for each of the 4 most common cancer types can easily be compared. 
``` {r exercise-3}
#The fraction of the genome altered for different cancer types is plotted and faceted by rows for each type.
msk_most_common %>% ggplot(aes(x = `Fraction Genome Altered`)) +
  geom_histogram() +
  facet_grid(rows = vars(`Cancer Type`), scales = "free_y")
#The x axis is kept fixed to allow for objective comparison of the mutation counts between cancer types, while the y-axis is freed as we are more interested in the proportion rather than the absolute counts within each type since we the number of cases is not equal. 
``` 

**Exercise 4 :** Boxplot of `Overall Survival (Months)` across various `Smoking History`. It is well established that smoking is a risk factor for various types of cancer, and thus this should also be reflected in our dataset. 
``` {r exercise-4}
#Rows with NA for the smoking history variable is excluded as we do not want to included NA as a factor for our boxplots. 
msk %>% filter(!is.na(`Smoking History`)) %>% ggplot(aes(x = `Smoking History`, y = `Overall Survival (Months)`)) +
  geom_boxplot()

#Since we do not stratify analysis by cancer type, we can use the initial dataset.
``` 

# 2.2 Exploration explained
The data exploration exercises performed both made the dataset more manageable to work with as well as offered us a basic understanding of our dataset to formulate research questions. 

1. Initially, the `msk` dataset contained `r nrow(msk)` cases of 59 different types of cancers. This woukd make it difficult to stratify for `Cancer Type` in our analysis, so the dataset was filtered to only contain the 4 most common types of cancer, which were: `r most_common_cancer_types`. This reduced the dataset to 4,716 cases, or `r as.integer(nrow(msk_most_common)/nrow(msk)*100)`% of the initial dataset.

2. Next, I wanted to see the density distribution of `Mutation Count` for different types of cancer. It is well established that different cancer types have their unique mutational characteristics, so it was reassuring to see this reflected in our data.

3. Similarly, I wanted to see the fraction of the genome altered differed for different types of cancer. Genomic instability is a hallmark of cancer, and also varies across different types. 

4. Finally, I looked at the overall survival for patients according to their `Smoking History`. Although it is known smoking is a risk factor for cancer, I wanted to see the effect on overall survival in our dataset. 

# 3.1 Research questions
Following our exploratory data analysis, I formulated the following 4 research questions: 

1. Which `Cancer Type` is the most deadly?

2. What variables are predictive of a patient's overall survival?

3. What are the genomic and mutational differences between primary and Metastasis samples?

4. What are the genomic and mutational consequences of a patient's smoking history?